---
title: "Tri-State Census Exploration: OH, KY, IN"
author: "Benedict Leonardi"
format: 
  html:
    theme: flatly
    toc: true
execute:
  warning: false
  message: false
---

## Overview

This document explores tract-level demographic data across Ohio, Kentucky, and Indiana. We use the `tidycensus` package to pull the data directly from the US Census Bureau, and `leaflet` to create an interactive map.

## Setup and Data Retrieval

First, we load our necessary libraries and fetch the American Community Survey (ACS) 5-year estimates. 

*Note: You must have a Census API key installed. If you haven't done this, run `census_api_key("YOUR_KEY_HERE", install = TRUE)` in your R console once.*

---
title: "Tri-State Census Exploration: OH, KY, IN"
author: "Benedict Leonardi"
format: 
  html:
    theme: flatly
    toc: true
execute:
  warning: false
  message: false
---

## Overview

This document explores tract-level demographic data across Ohio, Kentucky, and Indiana. We use the `tidycensus` package to pull the data and the `sf` package to calculate population and housing density per square mile.

## Setup and Data Retrieval

```{r setup-and-fetch}
#| label: fetch-data
library(tidycensus)
library(tidyverse)
library(sf)
library(leaflet)

# Define the variables
census_vars <- c(
  per_capita_income = "B19301_001",
  housing_units     = "B25001_001",
  population        = "B01003_001"
)

# Fetch the data
tri_state_data <- get_acs(
  geography = "cbg",
  variables = census_vars,
  state = c("OH", "KY", "IN"),
  year = 2022, 
  geometry = TRUE,
  output = "wide"
)

# Transform the coordinate reference system to WGS84 for Leaflet
map_data <- st_transform(tri_state_data, 4326)

# --- NEW: Calculate Area and Density ---
map_data <- map_data %>%
  mutate(
    # st_area() returns square meters. Divide by ~2.59 million to get square miles.
    area_sqmi = as.numeric(st_area(geometry)) / 2589988.11,
    
    # Calculate densities
    pop_density = populationE / area_sqmi,
    housing_density = housing_unitsE / area_sqmi
  )
```

## Interactive Map

Below is an interactive map. Use the layer control box in the top right corner to switch between Population, Income per Capita, and Housing Units.

```{r interactive-map}
#| label: interactive-map
#| out-width: "100%"
#| out-height: "600px"

# 1. Create color palettes
pal_pop <- colorNumeric("viridis", domain = map_data$populationE, na.color = "transparent")
pal_inc <- colorNumeric("magma", domain = map_data$per_capita_incomeE, na.color = "transparent")
pal_house <- colorNumeric("plasma", domain = map_data$housing_unitsE, na.color = "transparent")

# New palettes for density (using quantile binning can look better for highly skewed density data)
pal_pop_dens <- colorQuantile("viridis", domain = map_data$pop_density, n = 7, na.color = "transparent")
pal_house_dens <- colorQuantile("plasma", domain = map_data$housing_density, n = 7, na.color = "transparent")

# 2. Build the Leaflet map
leaflet(map_data) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  
  # --- Original Layers ---
  addPolygons(
    fillColor = ~pal_pop(populationE), color = "white", weight = 0.5, fillOpacity = 0.7,
    group = "Population (Count)",
    label = ~paste0(NAME, "<br/>Pop: ", scales::comma(populationE)) %>% lapply(htmltools::HTML)
  ) %>%
  addPolygons(
    fillColor = ~pal_inc(per_capita_incomeE), color = "white", weight = 0.5, fillOpacity = 0.7,
    group = "Income per Capita",
    label = ~paste0(NAME, "<br/>Income: ", scales::dollar(per_capita_incomeE)) %>% lapply(htmltools::HTML)
  ) %>%
  addPolygons(
    fillColor = ~pal_house(housing_unitsE), color = "white", weight = 0.5, fillOpacity = 0.7,
    group = "Housing Units (Count)",
    label = ~paste0(NAME, "<br/>Housing: ", scales::comma(housing_unitsE)) %>% lapply(htmltools::HTML)
  ) %>%
  
  # --- NEW: Density Layers ---
  addPolygons(
    fillColor = ~pal_pop_dens(pop_density), color = "white", weight = 0.5, fillOpacity = 0.7,
    group = "Population Density",
    label = ~paste0(NAME, "<br/>Pop Density: ", scales::comma(round(pop_density, 1)), " per sq mi") %>% lapply(htmltools::HTML)
  ) %>%
  addPolygons(
    fillColor = ~pal_house_dens(housing_density), color = "white", weight = 0.5, fillOpacity = 0.7,
    group = "Housing Density",
    label = ~paste0(NAME, "<br/>Housing Density: ", scales::comma(round(housing_density, 1)), " per sq mi") %>% lapply(htmltools::HTML)
  ) %>%
  
  # --- Dropdown Control ---
  addLayersControl(
    baseGroups = c("Population Density", "Housing Density", "Income per Capita", "Population (Count)", "Housing Units (Count)"),
    options = layersControlOptions(collapsed = FALSE)
  )
```

